S1: Extended Derivations for `A Data Driven Framework for Patient-Specific Neural Field Modelling'

### 2010-06-27 ltx/PLoS/LaTeX_Manuscript.tex
small edits to paper

The sensors were spaced on a regular square $14 \times 14$ grid and the sensor width was set to $\sigma^2_m = 0.81$. This provides a width of $1.5$~mm at half the maximum amplitude of the sensor kernel. The distance between sensors was set to $\Delta_y = 1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical for neurophysiological recordings.
 \begin{figure}
    	\begin{center}
     		\includegraphics{./Graph/ExperimentalFigure.pdf}
    	\end{center}
    	\caption{{\bf Example of the spatio-temporal properties of the model and synthetically generated data.} \textbf{(A)} The neural field at a single time instant. (\textbf{B}) Data generated by the full neural field model from the first 5 sensors. (\textbf{C}) Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of intracranial EEG.} 
 \label{fig:experimental design}
 \end{figure}
\subsection*{Spatial Frequency Analysis} 
Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency analysis was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FFTTrueEstimate}(A) shows the spatial frequency of the neural field. The cut-off frequency of the neural field is $\boldsymbol{\nu}_c \approx 0.24$~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of $2.05$~mm. This confirms that the separation of $\Delta_{y} = 1.5$~mm was sufficient to prevent problems associated with spatial aliasing.
\begin{figure}
\centering
\label{fig:FieldFFT}\includegraphics{./Graph/FFTField.pdf}
\caption{{\bf Spatial frequency analysis of the neural field}. (\textbf{A}) The average (over time) power in dB of the spatial frequency of the neural field. (\textbf{B}) The average (over time) power in dB of the spatial frequency of the reconstructed neural field from the basis function decomposition. The reconstructed field shows a narrower bandwidth as a result of the band limiting properties of the sensors and basis functions.}
\label{fig:FFTTrueEstimate}
\end{figure}

The spatial frequency of the observed field is shown in Figure~\ref{fig:ObservationFreqAnal}(C). The cut-off frequency is $\boldsymbol{\nu}_{cy} \approx 0.2$~Hz (-3~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:ObservationFreqAnal}(A). Using an oversampling parameter of $\rho_{\phi}=1$, the minimum distance between the field basis functions was calculated to be 2.5~mm to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). The distance between basis functions was set to $\Delta_{\phi}=2.5$~mm, which is sufficient to capture the significant dynamics according to our criteria.
Given the basis function separation, the basis function width parameter was set to $\sigma_{\phi}^2=2.5$~mm. This allowed a smooth representation of the field by the basis function decomposition up to a spatial frequency $\boldsymbol{\nu}_{c\phi} \approx 0.12$~Hz, as described by equation~\ref{eq:CutoffFromBasisFuncWidth}. The band-limiting effect is due to the spatial frequency response of the basis functions as seen in Figure~\ref{fig:ObservationFreqAnal}(B). The spatial frequency of the reconstructed field from the basis function decomposition is shown in Figure~\ref{fig:FFTTrueEstimate}(B). This figure illustrates the band-limiting effect of the basis function decomposition when compared to Figure~\ref{fig:FFTTrueEstimate}(A).
\begin{figure*}[!ht] 
	\begin{center}
		\includegraphics{./Graph/SensorBasisObsFrequencyResponse.pdf}	
	\end{center}
	\caption{{\bf Spatial frequency analysis}. The dashed line shows the cut-off frequency (-3 dB point). (\textbf{A}) Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the one-dimensional representation of the spectral characteristics. (\textbf{B}) The spatial frequency response of a one-dimensional model sensor. (\textbf{C}) The spatial frequency response of a one-dimensional neural field basis function, illustrating a narrower band width than the sensors.} 
	\label{fig:ObservationFreqAnal}
\end{figure*}
\subsection*{State and Parameter Estimates} 
\label{sec:state_and_param_results}
To demonstrate the performance of the state and parameter estimation algorithm, a Monte Carlo approach was employed where $100$ realisations of the data were generated. Each realisation consisted of $500$~ms of data (sampled at $1$~kHz) and the estimation was applied to the final $400$~ms, allowing the model's initial transients to die out. The initial state and parameters were unknown to the estimator. Eleven iterations of the estimation algorithm were used to ensure the parameters  converged. Figure~\ref{fig:ParametersConvergence} shows the rate of convergence, confirming the parameters converged, with the changes in parameters falling to less than $10^{-4}$ after six iterations.
\begin{figure}
        \centering
\includegraphics{./Graph/convergence.pdf}
\caption{{\bf Convergence of parameters.} The mean of the absolute error between true and estimated parameters averaged over 100 runs. All parameters converge after six iterations, where the mean difference between errors in consecutive iterations falls below $10^{-4}$.}
\label{fig:ParametersConvergence}
\end{figure}

Histograms of the parameter estimates obtained over the 100 realisations are shown in Figure~\ref{fig:Parameters}. The actual parameter values, the mean estimates, corresponding standard deviations and biases are given in Table~\ref{tab:Parameters estimates}. The biases show the percentage differences between the actual values and the mean parameter estimates. The mean estimates are in good concordance with the actual values.
\begin{figure}
    \centering
    \includegraphics{./Graph/ParametesDistributions.pdf}
  \caption{{\bf Histograms of the parameter estimates over 100
realisations}. The solid line shows the actual value of the parameters and the dotted line shows the mean of the estimated parameters. (\textbf{A}) The histogram of the $\xi$ estimates where $\xi=1-\zeta T_s $ and $\zeta$ is the reciprocal of the synaptic time constant. (\textbf{B}) The histogram of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. (\textbf{C}) The histogram of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$. (\textbf{D}) The histogram of the longer range excitatory connectivity kernel basis function parameter estimates, $\theta_2$.}
\label{fig:Parameters}
\end{figure}
The reconstructed connectivity kernels from the estimated parameters are shown in Figure~\ref{fig:KernelEstimates}, where the mean over the $100$ realisations is shown as a dotted line and the $95\%$ confidence interval is indicated by the grey shaded region. The figure verifies that the shape of the estimated connectivity kernel is in good accordance to the true kernel. The true kernel lies within the confidence interval; however, the mean estimated kernel is more negative than the actual kernel due to the small bias on the estimates of the inhibitory parameter, $\hat\theta_1$, as seen in Figure~\ref{fig:Parameters}(C).
\begin{figure}
    \centering
% \subfigure[][]{
\includegraphics{./Graph/KernelEstimate.pdf}%}
\caption{{\bf Cross-section of connectivity kernel estimate}. The actual kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the $95~\%$ confidence interval is shown by the shaded grey region.}
\label{fig:KernelEstimates}
\end{figure}
The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over iterations) of the root mean square error (MRMSE) (over space). The MRMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}, with the $95~\%$ confidence interval. At each time point the error is consistently around 0.5~mV, which is approximately $7.1$~\% of the field voltage range ($[-3.29,3.77]$~mV). The error in the reconstructed field is mostly due to the loss of high spatial frequency information from the basis function decomposition as illustrated in Figure~\ref{fig:FieldEstimate}. The figure shows a snapshot at a single time instant of the true neural field, the estimated field and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Figure~\ref{fig:FieldEstimate}(B) that the estimated field contains the basic structure of the true field (Figure~\ref{fig:FieldEstimate}(A)), but is missing the higher-frequency detail.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{{\bf Error in the field reconstruction}. The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region).} 
\label{fig:RMSE}
   \end{figure}

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
This research was funded in part by the Australian Research Council (Linkage Project LP0560684). The Bionic Ear Institute acknowledges the support it receives from the Victorian State Government through the Operational Infrastructure Support Program. Dean Freestone would like to thank the University of Melbourne's Scholarships Office for the Postgraduate Overseas Research Experience Scholarship and the Harold Mitchell Foundation for a traveling scholarship and supporting this research. In addition, Dean Freestone would like to thank Dr. Mark van Rossum for hosting him whilst visiting the University of Edinburgh where this work was undertaken.

>>>>

The sensors were spaced on a regular square $14 \times 14$ grid and the sensor width was set to $\sigma^2_m = 0.81$. This provides a width of 1.5~mm at half the maximum amplitude of the sensor kernel. The distance between sensors was set to $\Delta_y = 1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical for neurophysiological recordings.
 \begin{figure}
    	\begin{center}
     		\includegraphics{./Graph/ExperimentalFigure.pdf}
    	\end{center}
    	\caption{{\bf Example of the spatio-temporal properties of the model and synthetically generated data.} \textbf{(A)} The neural field at a single time instant. (\textbf{B}) Data generated by the full neural field model from the first 5 sensors. (\textbf{C}) Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of intracranial EEG.} 
 \label{fig:experimental design}
 \end{figure}
\subsection*{Spatial Frequency Analysis} 
Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency analysis was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FFTTrueEstimate}(A) shows the spatial frequency of the neural field. The cut-off frequency of the neural field is $\boldsymbol{\nu}_c \approx 0.24$~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of 2.05~mm. This confirms that the separation of $\Delta_{y} = 1.5$~mm was sufficient to prevent problems associated with spatial aliasing.
\begin{figure}
\centering
\label{fig:FieldFFT}\includegraphics{./Graph/FFTField.pdf}
\caption{{\bf Spatial frequency analysis of the neural field}. (\textbf{A}) The average (over time) power in dB of the spatial frequency of the neural field. (\textbf{B}) The average (over time) power in dB of the spatial frequency of the reconstructed neural field from the basis function decomposition. The reconstructed field shows a narrower bandwidth as a result of the band limiting properties of the sensors and basis functions.}
\label{fig:FFTTrueEstimate}
\end{figure}

The spatial frequency of the observed field is shown in Figure~\ref{fig:ObservationFreqAnal}(C). The cut-off frequency is $\boldsymbol{\nu}_{cy} \approx 0.2$~Hz (-3~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:ObservationFreqAnal}(A). Using an oversampling parameter of $\rho_{\phi}=1$, the minimum distance between the field basis functions was calculated to be 2.5~mm to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). The distance between basis functions was set to $\Delta_{\phi}=2.5$~mm, which is sufficient to capture the significant dynamics according to our criteria.
Given the basis function separation, the basis function width parameter was set to $\sigma_{\phi}^2=2.5$~mm. This allowed a smooth representation of the field by the basis function decomposition up to a spatial frequency $\boldsymbol{\nu}_{c\phi} \approx 0.12$~Hz, as described by equation~\ref{eq:CutoffFromBasisFuncWidth}. The band-limiting effect is due to the spatial frequency response of the basis functions as seen in Figure~\ref{fig:ObservationFreqAnal}(B). The spatial frequency of the reconstructed field from the basis function decomposition is shown in Figure~\ref{fig:FFTTrueEstimate}(B). This figure illustrates the band-limiting effect of the basis function decomposition when compared to Figure~\ref{fig:FFTTrueEstimate}(A).
\begin{figure*}[!ht] 
	\begin{center}
		\includegraphics{./Graph/SensorBasisObsFrequencyResponse.pdf}	
	\end{center}
	\caption{{\bf Spatial frequency analysis}. The dashed line shows the cut-off frequency (-3 dB point). (\textbf{A}) Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the one-dimensional representation of the spectral characteristics. (\textbf{B}) The spatial frequency response of a one-dimensional model sensor. (\textbf{C}) The spatial frequency response of a one-dimensional neural field basis function, illustrating a narrower band width than the sensors.} 
	\label{fig:ObservationFreqAnal}
\end{figure*}
\subsection*{State and Parameter Estimates} 
\label{sec:state_and_param_results}
To demonstrate the performance of the state and parameter estimation algorithm, a Monte Carlo approach was employed where 100 realisations of the data were generated. Each realisation consisted of 500~ms of data (sampled at $1$~kHz) and the estimation was applied to the final 400~ms, allowing the model's initial transients to die out. The initial state and parameters were unknown to the estimator. Eleven iterations of the estimation algorithm were used to ensure the parameters  converged. Figure~\ref{fig:ParametersConvergence} shows the rate of convergence, confirming the parameters converged, with the changes in parameters falling to less than $10^{-4}$ after six iterations.
\begin{figure}
        \centering
\includegraphics{./Graph/convergence.pdf}
\caption{{\bf Convergence of parameters.} The mean of the absolute error between true and estimated parameters averaged over 100 runs. All parameters converge after six iterations, where the mean difference between errors in consecutive iterations falls below $10^{-4}$.}
\label{fig:ParametersConvergence}
\end{figure}

Histograms of the parameter estimates obtained over the 100 realisations are shown in Figure~\ref{fig:Parameters}. The actual parameter values, the mean estimates, corresponding standard deviations and biases are given in Table~\ref{tab:Parameters estimates}. The biases show the percentage differences between the actual values and the mean parameter estimates. The mean estimates are in good concordance with the actual values.
\begin{figure}
    \centering
    \includegraphics{./Graph/ParametesDistributions.pdf}
  \caption{{\bf Histograms of the parameter estimates over 100
realisations}. The solid line shows the actual value of the parameters and the dotted line shows the mean of the estimated parameters. (\textbf{A}) The histogram of the $\xi$ estimates where $\xi=1-\zeta T_s $ and $\zeta$ is the reciprocal of the synaptic time constant. (\textbf{B}) The histogram of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. (\textbf{C}) The histogram of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$. (\textbf{D}) The histogram of the longer range excitatory connectivity kernel basis function parameter estimates, $\theta_2$.}
\label{fig:Parameters}
\end{figure}
The reconstructed connectivity kernels from the estimated parameters are shown in Figure~\ref{fig:KernelEstimates}, where the mean over the 100 realisations is shown as a dotted line and the 95\% confidence interval is indicated by the grey shaded region. The figure verifies that the shape of the estimated connectivity kernel is in good accordance to the true kernel. The true kernel lies within the confidence interval; however, the mean estimated kernel is more negative than the actual kernel due to the small bias on the estimates of the inhibitory parameter, $\hat\theta_1$, as seen in Figure~\ref{fig:Parameters}(C).
\begin{figure}
    \centering
% \subfigure[][]{
\includegraphics{./Graph/KernelEstimate.pdf}%}
\caption{{\bf Cross-section of connectivity kernel estimate}. The actual kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the 95~\% confidence interval is shown by the shaded grey region.}
\label{fig:KernelEstimates}
\end{figure}
The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over iterations) of the root mean square error (MRMSE) (over space). The MRMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}, with the $95~\%$ confidence interval. At each time point the error is consistently around 0.5~mV, which is approximately 7.1~\% of the field voltage range ([-3.29,3.77]~mV). The error in the reconstructed field is mostly due to the loss of high spatial frequency information from the basis function decomposition as illustrated in Figure~\ref{fig:FieldEstimate}. The figure shows a snapshot at a single time instant of the true neural field, the estimated field and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Figure~\ref{fig:FieldEstimate}(B) that the estimated field contains the basic structure of the true field (Figure~\ref{fig:FieldEstimate}(A)), but is missing the higher-frequency detail.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{{\bf Error in the field reconstruction}. The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region).} 
\label{fig:RMSE}
   \end{figure}

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
Dean Freestone would like to thank Dr. Mark van Rossum for hosting him whilst visiting the University of Edinburgh where this work was undertaken.

% This research was funded in part by the Australian Research Council (Linkage Project LP0560684). The Bionic Ear Institute acknowledges the support it receives from the Victorian State Government through the Operational Infrastructure Support Program. Dean Freestone would like to thank the University of Melbourne's Scholarships Office for the Postgraduate Overseas Research Experience Scholarship and the Harold Mitchell Foundation for a traveling scholarship and supporting this research.
